---
title: "Amanda"
author: "Maximo Nores"
format: pdf
editor: visual
---

## 

## Load Packages

```{r}
library(tidyverse)
library(nlraa)
library(nlme)
library(skimr)
library(dplyr)
library(lubridate)

library(readxl)
```

## Load Data

```{r}
amanda <- read_excel("../data/compiled.xlsx")
amanda <- na.omit(amanda)
```

## EDT

```{r}
skim(amanda)
```

```{r}
hist(amanda$Yld)
boxplot(amanda$Yld)
```

## 

```{r}
amanda %>%
  ggplot(aes(x= Test, y= Yld))+
  geom_boxplot()+
  theme_minimal()
```

Transform to factors

```{r}
amanda$Year <- as.factor(amanda$Year)
#amanda$Variety <- as.factor(amanda$Variety)
amanda$Test <- as.factor(amanda$Test)
amanda$Block <- as.factor(amanda$Block)
amanda$Loc <- as.factor(amanda$Loc)
amanda$Brand <- as.factor(amanda$Brand)
amanda$Rep <- as.factor(amanda$Rep)
str(amanda)
```

# Fitting models

```{r}
library(lme4)
library(lmerTest)
```

## Mod 1

No interaction between test and loc

```{r}
model1 <- lmer(Yld ~ Test + Loc + (1 | Variety) + (1 | Block) + (1|Rep), data = amanda)
```

```{r}
summary(model1)
anova(model1)

```

## Mod 2

model 2 includes year as a random effect

```{r}
model2 <- lmer(Yld ~ Test + Loc + (1 | Variety) + (1 | Block) + (1|Rep)+ (1|Year), data = amanda)
```

## Model 3

model 3 includes the interaction between test and loc.

```{r}
model3 <- lmer(Yld ~ Test * Loc + (1 | Variety) + (1 | Block) + (1|Rep)+ (1|Year), data = amanda)
```

## Model 2

Model 2 adds Brand as a random effect

```{r}
model4 <- lmer(Yld ~ Test * Loc + (1 | Variety) + (1 | Block) + (1|Rep)+ (1|Year) + (1|Brand), data = amanda)
```

## Model comaprison

```{r}
anova(model1,model2, model3, model4)
```

Model 4 is the best one, has the lowest BIC and AIC.

### Checking Residuals

```{r}
qqnorm(residuals(model4))
qqline(residuals(model4), col = "red")

```

```{r}
plot(fitted(model1), residuals(model1),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)
```

```{r}
library(emmeans)
emmeans_test <- emmeans(model4, ~ Test|Loc)
pairs(emmeans_test, adjust = "tukey")
```

## Plasticity

phenotipic plasticity (PPIK) = variance of yield for crm across env (VTIP)/ variance of all crm across env (vtk)

```{r}
amanda$env <- interaction(amanda$Loc, amanda$Year, drop= TRUE)

amanda$predicted_yield <- predict(model4, newdata= amanda ,level= 0)
```

mean predicted yield by environment

```{r}
mean_pred <- amanda %>%
  group_by(Test, env) %>%
  summarise(mean_pred_yield = mean(predicted_yield), .groups = 'drop')
```

Variance of predicted yield by Test

### VTIP

```{r}
plasticity_df <- mean_pred %>%
  group_by(Test) %>%
  summarise(variance_across_env = var(mean_pred_yield), .groups = 'drop')
```

### Total variance= VTK

```{r}
total_variance <- var(mean_pred$mean_pred_yield)
```

### Plasticity PPIK

```{r}
plasticity_df <- plasticity_df %>%
  mutate(plasticity_index = variance_across_env / total_variance)

print(plasticity_df)
```

```{r}
ggplot(plasticity_df, aes(x = Test, y = plasticity_index, fill = Test)) +
  geom_col() +
  labs(title = "Phenotypic Plasticity by CRM Group",
       x = "CRM ",
       y = "Plasticity Index ") +
  theme_minimal()
```

```{r}
mean_yield_df <- amanda %>%
  group_by(Test) %>%
  summarise(mean_predicted_yield = mean(predicted_yield), .groups = "drop")

combined_df <- left_join(plasticity_df, mean_yield_df, by = "Test")

ggplot(combined_df, aes(x = plasticity_index, y = mean_predicted_yield, label = Test)) +
  geom_point() +
  geom_text(vjust = -0.5, size = 3) +
  labs(title = "Stability (Plasticity) vs Yield",
       x = "Plasticity Index (Lower = More Stable)",
       y = "Mean Predicted Yield") +
  theme_minimal()
```

Testing correlation between yield and plasticity

```{r}
cor.test(combined_df$plasticity_index, combined_df$mean_predicted_yield)
```

As plasticity increases, yield tends to decrease.
