---
title: "mixed_effect_models"
output: pdf_document
---

## Load packages

```{r}
library(tidyverse)
library(nlraa)
library(nlme)
library(skimr)
library(dplyr)
library(lubridate)
```

## Load data

```{r}
variety.testing <- read.csv("2025-maximo-and-elvis/raw-data/Corn data organized-10-16-2024.csv")
```

## Exploratory analysis


```{r}
variety.testing <- variety.testing %>%
  mutate(CRM_group = case_when(
    #CRM %in% c(103, 107, 108) ~ "Group_1",
    CRM %in% c(110, 111, 112, 113) ~ "Group_2",
    CRM %in% c(114, 115, 116, 117) ~ "Group_3",
    CRM %in% c(118, 119, 120, 121) ~ "Group_4"))

```

```{r}
fac.cols <- c('Location', 'Hybrid', 'Year', 'row_spacing', 'CRM_group')
variety.testing[c(fac.cols)] <- lapply(variety.testing[fac.cols], function(x) as.factor(x))
str(variety.testing)
```

Eliminating NA's

```{r}
variety.testing <- na.omit(variety.testing)
```

## Fitting Models

```{r}
mod6_loc <- lme(yield ~ CRM_group + SowingdateDOY * Location + CRM_group : Location + CRM_group:SowingdateDOY:Location,
                random = ~ 1| Year,
                weights = varIdent(form= ~ 1 | Location),
                data = variety.testing)
anova(mod6_loc)
```

# Changing approach 

## Selecting upper 10% of yields 

## Load packages

```{r}
library(tidyverse)
library(nlraa)
library(nlme)
library(skimr)
library(dplyr)
library(lubridate)
library(MuMIn)
library(emmeans)
```

##p1
```{r}
var.test <- read.csv("input/Corn data organized1.csv")
```


```{r}
var.test <- var.test %>%
  mutate(crm_group = case_when(
    crm %in% c(103, 107, 108) ~ "Group_1",
    crm %in% c(110, 111, 112, 113) ~ "Group_2",
    crm %in% c(114, 115, 116, 117) ~ "Group_3",
    crm %in% c(118, 119, 120, 121) ~ "Group_4"))
var.test <- var.test %>%
  
  select(location, year, hybrid, sowingdateDOY, yield, week, crm_group,mean_maxt, mean_mint, cum_radn,cum_rain,above_32, above_35,region, april, may, june, july, august, sept)

var.test <- var.test %>%
  mutate(year = as.factor(year),
    location = as.factor(location),
    sowingdateDOY = as.factor(sowingdateDOY),
    crm_group = as.factor(crm_group),
    yield= as.numeric(yield))%>%
  na.omit(var.test)  
#var.test <- var.test %>%
#  mutate(SowingdateDOY2 = SowingdateDOY^2)
#var.test <- var.test %>%
#  group_by(week) %>%
#  mutate(top10 = Yield >= quantile(Yield, 0.90, na.rm = TRUE)) %>%
#  ungroup()
```

```{r}
var.test.se <- var.test %>% 
  filter(region== "SouthEast")
```


```{r}
weather.se <- lme(yield ~ above_35 + mean_mint + cum_rain + cum_radn,
               random = ~ 1 |year/location,
               #weights = varIdent(form = ~1 | year),
  data = var.test.se)
  
r.squaredGLMM(weather.se)
```

```{r}
var.test.ce <- var.test %>% 
  filter(region== "CenterEast")
```

```{r}
weather.ce <- lme(yield ~ above_35 + mean_mint + cum_rain + cum_radn ,
               random = ~ 1 |year/location,
               #weights = varIdent(form = ~1 | year),
  data = var.test.ce)
  
r.squaredGLMM(weather.ce)
```


```{r}
var.test.ne <- var.test %>% 
  filter(region== "NorthEast") %>% 
  na.omit()
```

```{r}
weather.ne <- lme(yield ~ above_35 + mean_mint + cum_rain + cum_radn,
               random = ~ 1 |year/location,
               #weights = varIdent(form = ~1 | year),
  data = var.test.ne)
  
r.squaredGLMM(weather.ne)
```

```{r}
weather <- lme(yield ~ above_35 + mean_mint + cum_rain + cum_radn,
               random = ~ 1 |year/location,
               #weights = varIdent(form = ~1 | year),
  data = var.test)
r.squaredGLMM(weather)
```


```{r}
var.top10 %>%
  ggplot(aes(x=SowingdateDOY, y= Yield))+
  geom_point()+
  geom_smooth(method = "lm", formula = y ~ poly(x, 2))+
  scale_x_continuous(breaks = seq(80, 145, by=5))+
  theme_minimal()
```
Using entire dataset to fit the model. Then select upper 10%
```{r}
top.mod <- lme(
  Yield ~ SowingdateDOY + CRM,
  random = ~1 | Year/Location,
  #weights = varIdent(form= ~ 1 | Year),
  data = var.test)
top.mod.q <- lme(
  Yield ~ SowingdateDOY + SowingdateDOY2 + CRM ,
  random = ~1 | Year/Location,
  weights = varIdent(form = ~1 | Location),
  data = var.test)

anova(top.mod.q)
IC_tab(top.mod, top.mod.q)
plot(top.mod.q)
qqnorm(resid(top.mod.q)); qqline(resid(top.mod.q))
hist(resid(top.mod.q), breaks = 30)
```

Quadratic model is better
```{r}
top.mod1.q <- lme(
  Yield ~ SowingdateDOY + SowingdateDOY2 + crm.f,
  random = ~1 | Year/Location,
  weights = varIdent(form = ~1 | Location),
  data = var.test)
anova(top.mod1.q)
IC_tab(top.mod.q, top.mod1.q)
plot(top.mod1.q)
qqnorm(resid(top.mod1.q)); qqline(resid(top.mod1.q))
hist(resid(top.mod1.q), breaks = 30)
```

Grabbing the upper 10% of yields from the model

```{r}
var.test <- var.test %>%
  group_by(week) %>%
  mutate(top10_fit = Yield >= quantile(Yield, 0.90, na.rm = TRUE)) %>%
  ungroup()
```


```{r}
mod_top_line <- lm(Yield ~ SowingdateDOY + SowingdateDOY2, 
                   data = filter(var.test, top10_fit == TRUE))

```

```{r}
newdata <- data.frame(SowingdateDOY = seq(min(var.test$SowingdateDOY),
                                          max(var.test$SowingdateDOY),
                                          length.out = 100))
newdata$SowingdateDOY2 <- newdata$SowingdateDOY^2
newdata$pred <- predict(mod_top_line, newdata)

```


```{r}
p1 <- ggplot(var.test, aes(x= SowingdateDOY, y= Yield))+
  geom_point(alpha= 0.1)+
  geom_point(data= filter(var.test, top10_fit == TRUE),
             aes(x= SowingdateDOY, y= Yield))+
  geom_line(data= newdata,aes(x= SowingdateDOY, y= pred),
            color= "red", size= 1.2)+
  scale_x_continuous(breaks= seq(80,150,by= 5))+
  scale_y_continuous(breaks= seq(150,350,by= 25))+
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA))

ggsave("output/sowingdate_yield_plot.tiff", plot = p1,
       dpi = 600, width = 8, height = 6, units = "in", compression = "lzw", bg= "white") 
```

##p2


```{r}
data.org <- read_csv("input/Corn data organized2.csv",   na = c("#N/A", "NA")) 
data.org$env <- interaction(data.org$location, data.org$year, drop= TRUE)
```

```{r}
data.org <- data.org %>%
  mutate(crm_group = case_when(
    #crm %in% c(103, 107, 108) ~ "Group_1",
    crm %in% c(110, 111, 112, 113) ~ "Group_2",
    crm %in% c(114, 115, 116, 117) ~ "Group_3",
    crm %in% c(118, 119, 120, 121) ~ "Group_4")) %>%
  select(location, year, hybrid, crm, yieldkg.ha, crm_group)%>%
  filter(year != 2020)
  #mutate(sowing.doy2 = sowing.doy^2)

data.org <- na.omit(data.org)
data.org <- data.org %>%
  mutate(
    year= factor(year),
    location = as.factor(location),
    yieldkg.ha = as.numeric(yieldkg.ha),
    crm_group = as.factor(crm_group),
    hybrid = factor(hybrid))
```



```{r}
mod_emm1.1 <- lme(
  yieldkg.ha ~  crm_group * year,
  random = ~1 | location,
  #weights = varIdent(form = ~1 | location),
  data = data.org)
```

```{r}
plot(mod_emm1.1, which = 1)
qqnorm(resid(mod_emm1.1))
qqline(resid(mod_emm1.1), col = "red", lwd = 2)
```

```{r}
emm1.1 <- emmeans(mod_emm1.1, ~ crm_group | year)
emm_df1.1 <- as.data.frame(emm1.1)
```

```{r}
p2 <- ggplot(emm_df1.1, aes(x = year, y = emmean, color = crm_group,
                         group = crm_group))+
  geom_point(size = 3) +
  geom_line(size = 1.2) +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.2) +
  scale_color_manual(
    values = c("#1b9e77","#fc8d62","#1f78b4","#525252")) +
  labs(
    x = "Year",
    y = "Adjusted Yield (kg ha⁻¹)",
    color = "CRM Group") +
  theme_minimal()+scale_color_manual(
  values = c("#1b9e77","#fc8d62","#1f78b4","#525252"),
  labels = c(
    "Group 2 (110d–113d)",
    "Group 3 (114d–117d)",
    "Group 4 (118d–121d)")) +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA))
ggsave("output/emmeans.year.crm_group.tiff", plot = p2,
       dpi = 600, width = 8, height = 6, units = "in", compression = "lzw", bg= "white")
```

## emmeans
```{r}
library(dplyr)
library(tidyverse)
library(nlme)
library(nlraa)
library(r2glmm)
library(janitor)
library(biometryassist)
library(emmeans)
```

##p4

```{r}
data.org <- read_csv("input/Corn data organized2.csv", na = c("#N/A", "NA")) %>%
  mutate(
    crm_group = case_when(
      crm %in% c(103, 107, 108) ~ "Group_1",
      crm %in% c(110, 111, 112, 113) ~ "Group_2",
      crm %in% c(114, 115, 116, 117) ~ "Group_3",
      crm %in% c(118, 119, 120, 121) ~ "Group_4"),
    location = factor(location),
    year = factor(year),
    hybrid = factor(hybrid),
    region = factor(region),
    yieldkg.ha = as.numeric(yieldkg.ha)) %>%
  select(location, year, hybrid, crm_group, region, sowing.doy, yieldkg.ha) %>%
  na.omit() %>%
  mutate(env = interaction(location, year, drop = TRUE))
```

```{r}
mod_simple <- lme(yieldkg.ha ~ env, random = ~1 | hybrid, data = data.org)

emm_df2 <- as.data.frame(emmeans(mod_simple, ~ env)) %>%
  separate(env, into = c("location", "year"), sep = "\\.", remove = FALSE) %>%
  mutate(
    location = factor(location, levels = c("Harrisburg","Keiser",
                                           "Marianna",
                                           "Des_Arc","Rohwer","Stuttgart")),
    year = factor(year))
```

```{r}
loc_labels <- c(
  Harrisburg = "NorthEast 1",
  Keiser     = "NorthEast 2",
  Marianna   = "CenterEast 1",
  Des_Arc   = "CenterEast 2",
  Rohwer     = "SouthEast 1",
  Stuttgart  = "SouthEast 2")
```

```{r}
data.org <- data.org %>%
  mutate(location = factor(location, 
                           levels = c("Harrisburg","Keiser", # NorthEast
                                      "Marianna","Des_Arc",                # CenterEast
                                      "Rohwer","Stuttgart")))      # SouthEast
```


```{r}
p4 <- ggplot(data.org, aes(x = location, y = yieldkg.ha, color = location)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(data = emm_df2, 
             aes(x = location, y = emmean, group = year, color = location),
             position = position_dodge(width = 0.6), size = 3, inherit.aes = FALSE) +
  geom_errorbar(data = emm_df2,
                aes(x = location, ymin = lower.CL, ymax = upper.CL, group = year, color = location),
                width = 0.15, position = position_dodge(width = 0.6), inherit.aes = FALSE) +
  scale_x_discrete(labels = loc_labels) +
  guides(color = "none") +
  labs(
  x = "Region",
  y = "Yield (kg ha⁻¹)",
  caption = paste(
    "NorthEast 1 = Harrisburg;",
    "NorthEast 2 = Keiser;",
    "CenterEast 1 = Marianna;",
    "CenterEast 2 = Des Arc;",
    "SouthEast 1 = Rohwer;",
    "SouthEast 2 = Stuttgart")) +
  theme_minimal() +
  theme(
    axis.text.x   = element_text(size = 12),
    axis.text.y   = element_text(size = 12),
    axis.title    = element_text(size = 14), 
    plot.caption  = element_text(hjust = 0,  size = 10),
    plot.background  = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA))
ggsave("output/boxplots.tiff", plot = p4,
       dpi = 600, width = 11, height = 6, units = "in", compression = "lzw")
```

```{r}
data.org <- read_csv("input/Corn data organized2.csv",   na = c("#N/A", "NA"))
data.org$env <- interaction(data.org$location, data.org$year, drop= TRUE)
```

```{r}
data.org <- data.org %>%
  mutate(crm_group = case_when(
    crm %in% c(103, 107, 108) ~ "Group_1",
    crm %in% c(110, 111, 112, 113) ~ "Group_2",
    crm %in% c(114, 115, 116, 117) ~ "Group_3",
    crm %in% c(118, 119, 120, 121) ~ "Group_4")) %>%
  select(location, year, hybrid,crm,sowing.doy, yield,yieldkg.ha,
         crm_group,env,region) %>%
  #filter(year != 2020)
  mutate(sowing.doy2 = sowing.doy^2)

data.org <- na.omit(data.org)
data.org <- data.org %>%
  group_by(year) %>% 
  #mutate(unity = 1:length(year)) %>%
  ungroup() %>% 
  mutate(
    location = as.factor(location),
    yieldkg.ha = as.numeric(yieldkg.ha),
    crm_group = as.factor(crm_group),
    region= factor(region),
    env= factor(env))
```

```{r}
mod_emm2.1 <- lme(yield ~ env + sowing.doy + sowing.doy2 * crm_group,
  random = ~1 | hybrid,
 weights = varIdent(form = ~1 | location),
  data = data.org)
summary(mod_emm2.1)
```

mod_emm2 <- lme(yield ~ sowing.doy + sowing.doy2 * crm_group + cum_rain + cum_radn + mean_maxt + year,
  random = ~1 | hybrid,
 weights = varIdent(form = ~1 | location),
  data = data.org)
summary(mod_emm2)

```{r}
plot(mod_emm2.1, which = 1)
qqnorm(resid(mod_emm2.1))
qqline(resid(mod_emm2.1), col = "red", lwd = 2)
```

#emmeans
```{r}
emm_mod2 <- emmeans(mod_emm2.1, ~ env)
summary(emm_mod2)
emm_df2 <- as.data.frame(emmeans(mod_emm2.1, ~ env))
```

```{r}
emm <- emmeans(mod_emm2, ~ crm_group | year)
emm_df <- as.data.frame(emm)
```


```{r}
emm_df2 <- emm_df2 |>
  separate(env, into = c("location", "year"), sep = "\\.", remove = FALSE)
loc_order <- c("Harrisburg", "Keiser", "Marianna", "Bellfarm", "Rohwer", "Stuttgart")
data.org$location  <- factor(data.org$location,  levels = loc_order)
emm_df2$location   <- factor(emm_df2$location,   levels = loc_order)
emm_df2$year       <- factor(emm_df2$year)

```


```{r}
loc_labels <- c(
  Harrisburg = "NorthEast",
  Keiser     = "NorthEast",
  Marianna   = "CenterEast",
  Bellfarm   = "CenterEast",
  Rohwer     = "SouthEast",
  Stuttgart  = "SouthEast")
```

#Yield across locations
```{r}
 p3 <- ggplot(data.org, aes(x = location, y = yield, color = location)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(
    data = emm_df2,
    aes(x = location, y = emmean, group = year, color = location),
    position = position_dodge(width = 0.6),
    size = 3,
    inherit.aes = FALSE) +
  geom_errorbar(data = emm_df2,
    aes(x = location, ymin = lower.CL, ymax = upper.CL,
        group = year, color = location),
    width = 0.15,
    position = position_dodge(width = 0.6),
    inherit.aes = FALSE) +
  scale_x_discrete(labels = loc_labels) +
  guides(color = "none") +
  labs(x = "Region", y = "Yield",
       title = "Observed vs. Model-Estimated Corn Yields by Region") +
  theme_minimal()+
  theme(plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA))
ggsave("output/Observed vs. Model-Estimated Corn Yields across Locations.tiff", plot = p3,
       dpi = 600, width = 8, height = 6, units = "in", compression = "lzw")
```


```{r}
ggplot(data.org, aes(x = location, y = yield, color = location)) +
  geom_boxplot() +
  geom_jitter(width = 0.3, alpha = 0.2) +
  theme_minimal()
```
#p4

```{r}
data.org <- na.omit(data.org)
data.org <- data.org %>% mutate(week = as.character(week))

week_levels <- c(
  "march 23- march 29",
  "march 30- april 5",
  "april 6- april 12",
  "april 13- april 19",
  "april 20-april 26",
  "april 27- may 3",
  "may 11- may 17",
  "may 18- may 24")
wk <- data.org %>%
  group_by(week)%>%
  summarise(
    n          = sum(!is.na(Yield)),
    mean_yield = mean(Yield, na.rm = TRUE),
    sd         = ifelse(n > 1, sd(Yield, na.rm = TRUE), NA_real_),
    se         = ifelse(n > 1, sd / sqrt(n), NA_real_),
    .groups = "drop") %>%
  mutate(week = factor(week, levels = week_levels))

```

```{r}
ggplot(wk, aes(x = week, y = mean_yield)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_yield - 1.96*se,
                    ymax = mean_yield + 1.96*se),
                width = 0.2) +
  scale_y_continuous(breaks = seq(20,260,by= 20))+
  theme_minimal()
```
#p5

```{r}
ggplot(data.org, aes(x = week, y = yield)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() 
```

#p6

```{r}
mod <- lme(yield ~ crm_group + week + above_35 + cum_rain,
  random = ~1 | year/location,
  data = data.org)
emm_df3 <- emmeans(mod, ~ crm_group) %>%
  as.data.frame()

ggplot(emm_df3, aes(x = crm_group, y = emmean, fill = crm_group)) +
  geom_col() +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  labs(x = "CRM group", y = "Adjusted Yield") +
  theme_minimal()
anova(mod)
pairs(emmeans(mod, ~ crm_group))

```

```{r}
crm_levels <- c("Group_1","Group_2","Group_3","Group_4")

crm <- data.org %>%
  group_by(CRM_group) %>%
  summarise(
    n          = sum(!is.na(Yield)),
    mean_yield = mean(Yield, na.rm = TRUE),
    sd         = ifelse(n > 1, sd(Yield, na.rm = TRUE), NA_real_),
    se         = ifelse(n > 1, sd / sqrt(n), NA_real_),
    .groups = "drop") %>%
  mutate(CRM_group = factor(CRM_group, levels = crm_levels))

ggplot(crm, aes(x = CRM_group, y = mean_yield)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_yield - 1.96 * se,
                    ymax = mean_yield + 1.96 * se),
                width = 0.2) +
  scale_y_continuous(breaks = seq(20, 260, by = 20)) +
  labs(x = "CRM group", y = "Mean yield") +
  theme_minimal()
```

```{r}
ggplot(data.org, aes(x = CRM_group, y = Yield)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() 
```


###weather

```{r}
weather <- read.csv("input/weather_monthly_summary.csv")

```

```{r}
weather_summary <- weather %>%
  mutate(month = factor(month, levels = month.name)) %>%
  group_by(location, month) %>%
  summarise(
    mean_avg_t = mean(avg_t, na.rm = TRUE),
    mean_mint = mean(mint,na.rm= TRUE),
    mean_cum_rain = mean(cum_rain, na.rm = TRUE),
    mean_precipitation = mean(precipitation,na.rm = TRUE),
    .groups = "drop")

```

```{r}
loc_labels <- c(
  Des_Arc         = "CenterEast (Des Arc)",
  harrisburg_neerec = "NorthEast (Harrisburg)",
  keiser_nere       = "NorthEast (Keiser)",
  marianna          = "CenterEast (Marianna)",
  rohwer            = "SouthEast (Rohwer)",
  stuttgart         = "SouthEast (Stuttgart)")
weather_summary <- weather_summary %>%
  mutate(region = loc_labels[location])
```


#p7: average temperature

```{r}
 weather_summary %>%
  drop_na() %>% 
  ggplot(aes(x = month, y = mean_avg_t, color = region, group = region)) +
  geom_line(size = 1.1, linetype = "dashed") +
  geom_point(size = 3, stroke = 0.8) +
  scale_y_continuous(breaks = seq(10, 30, by = 2)) +
  scale_x_discrete(
    limits = c("April", "May", "June", "July", "August", "September"),
    labels = c("Apr", "May", "Jun", "Jul", "Aug", "Sep")) +
  scale_color_brewer(
    palette = "Dark2",
    name   = "Region") +
  labs(
    x = "Month",
    y = "Mean Avg Temp (°C)",
    title = "Monthly Average Temperature by Region") +
  theme_minimal(base_size = 14) 
  
ggsave("output/avg_temperature.tiff", plot = avg_temperature,
       dpi = 600, width = 8, height = 6, units = "in", compression = "lzw")
```

## Average rainfall
```{r}
 avg_rainfall <- weather_summary %>%
  filter(!(region == "NorthEast (Harrisburg)" & month == "September")) %>%
  ggplot(aes(x = month, y = mean_precipitation, color = region,
             group = region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_y_continuous(breaks = seq(20, 200, by = 20)) +
  labs(
    x = "Month",
    y = "Mean cumulative rain (mm)",
    color = NULL) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA))
ggsave("output/avg_rainfall.tiff", plot = avg_rainfall,
       dpi = 600, width = 8, height = 6, units = "in", compression = "lzw", background = "white")
```

```{r}
minimumT <- weather_summary %>%
 # filter(!(region == "NorthEast (Harrisburg)" & month == "September")) %>%
  ggplot(aes(x = month, y = mean_mint, color = region,
             group = region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_y_continuous(breaks = seq(10, 24, by = 2)) +
  labs(
    x = "Month",
    y = "Mean minimum Temperature (°C)",
    color = NULL) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA))
ggsave("output/NT.across.months.tiff", plot = minimumT,
       dpi = 600, width = 8, height = 6, units = "in", compression = "lzw", background = "white")
```




##Environmental model
#All locations together

```{r}
data.org <- read.csv("input/Corn data organized2.csv")
data.org$Yield <- as.numeric(data.org$Yield)
```


```{r}
data.org <- data.org %>%
  mutate(CRM_group = case_when(
    CRM %in% c(103, 107, 108) ~ "Group_1",
    CRM %in% c(110, 111, 112, 113) ~ "Group_2",
    CRM %in% c(114, 115, 116, 117) ~ "Group_3",
    CRM %in% c(118, 119, 120, 121) ~ "Group_4"))
data.org <- data.org %>%
  select(Location, Year, Hybrid, SowingdateDOY, CRM, Yield, week, CRM_group,
         mean_maxt, mean_mint, cum_radn,cum_rain,above_32, above_35,week,region)
data.org <- data.org %>%
  mutate(SowingdateDOY2 = SowingdateDOY^2)

data.org <- na.omit(data.org)
data.org <- data.org %>%
  mutate(top10 = Yield >= quantile(Yield, 0.90, na.rm = TRUE)) %>%
  ungroup()
data.org <- data.org %>%
  mutate(
    Year = as.factor(Year),
    Location = as.factor(Location),
    Yield = as.numeric(Yield),
    sowingdate.f = as.factor(SowingdateDOY),
    crm.f = as.factor(CRM_group))
```


check correlation between variables
```{r}
pairs(data.org[, c("yield", "mean_maxt", "mean_mint", "cum_radn", "cum_rain", "above_32", "above_35", "SowingdateDOY")])
```


```{r}
mod_emm1 <- lme(
  Yield ~  region + SowingdateDOY * CRM_group +
    above_35 + cum_rain + cum_radn + Year,
  random = ~1 | Location,
  weights = varIdent(form = ~1 | region),
  data = data.org)
anova(mod_emm1)
mod_emm1.2 <- lme(
  Yield ~  Location + SowingdateDOY * CRM_group +
    above_35 + cum_rain + cum_radn + Year,
  random = ~1 | Hybrid,
  weights = varIdent(form = ~1 | Location),
  data = data.org)
anova(mod_emm1.2)

```
adding the effect of planting date as quadratic
```{r}
mod_emm2 <- lme(
  Yield ~  Location + SowingdateDOY + SowingdateDOY2 * CRM_group +
    above_35 + cum_rain + cum_radn + Year,
  random = ~1 | Hybrid,
  weights = varIdent(form = ~1 | Location),
  data = data.org)
anova(mod_emm2)

```

```{r}
IC_tab(mod_emm1, mod_emm2)
plot(mod_emm2)
qqnorm(resid(mod_emm2)); qqline(resid(mod_emm2))
hist(resid(mod_emm2), breaks = 30)
```

```{r}
library(r2glmm)
```

```{r}
partial_r2 <- r2beta(mod_emm2, method = "nsj", data = data.org)
print(partial_r2)
```

```{r}
data.org <- data.org %>%
  mutate(Location = str_trim(Location)) %>%           
  filter(tolower(Location) != "harrisburg") 
by_location <- split(data.org, data.org$Location)

results <- map_df(by_location, function(df) {
  mod <- lme(
    Yield ~ SowingdateDOY + SowingdateDOY2 * CRM_group +
      above_35 + cum_rain + cum_radn,
    random = ~1 | Year/Hybrid,
    data = df)
  
  r2s <- r2beta(mod, method = "nsj", data = df)
  r2s$Location <- unique(df$Location)
  return(r2s)
})
```


```{r}
mod_rs <- lme(
  Yield ~ SowingdateDOY + SowingdateDOY2 * CRM_group +
    above_35 + cum_rain + cum_radn,
  random = ~ SowingdateDOY | Location,
  weights = varIdent(form = ~1 | Location),
  data = data.org)
mod_rs2 <- lme(
  Yield ~  Location + SowingdateDOY + SowingdateDOY2 * CRM_group +
    above_35 + cum_rain + cum_radn + Year,
  random = ~1 | Hybrid,
  weights = varIdent(form = ~1 | Location),
  data = data.org)

IC_tab(mod_rs, mod_rs2)
```

```{r}
partial_r2 <- r2beta(mod_rs2, method = "nsj", data = data.org)
print(mod_rs2)
```

```{r}
partial_r2 %>%
  mutate(Group = case_when(
    grepl("CRM_group|Hybrid", Effect) ~ "Genetics",
    grepl("SowingdateDOY|:CRM_group", Effect) ~ "Management",
    grepl("above_35|cum_rain|cum_radn|Year|Location", Effect) ~ "Environment",
    TRUE ~ "Other")) %>%
  group_by(Group) %>%
  summarise(Group_R2 = sum(Rsq)) %>%
  arrange(desc(Group_R2))

```

# By Region

```{r}
library(r2glmm)
```


```{r}
data.org1 <- read.csv("input/Corn data organized2.csv")
data.org1$Yield <- as.numeric(data.org1$Yield)
```


```{r}
data.org1 <- data.org1 %>%
  mutate(CRM_group = case_when(
    CRM %in% c(103, 107, 108) ~ "Group_1",
    CRM %in% c(110, 111, 112, 113) ~ "Group_2",
    CRM %in% c(114, 115, 116, 117) ~ "Group_3",
    CRM %in% c(118, 119, 120, 121) ~ "Group_4"))
data.org1 <- data.org1 %>%
  select(Location, Year, Hybrid, SowingdateDOY, CRM, Yield, week, CRM_group,
         mean_maxt, mean_mint, cum_radn,cum_rain,above_32, above_35,week,region)
data.org1 <- data.org1 %>%
  mutate(SowingdateDOY2 = SowingdateDOY^2)

data.org1 <- na.omit(data.org1)
data.org1 <- data.org1 %>%
  mutate(top10 = Yield >= quantile(Yield, 0.90, na.rm = TRUE)) %>%
  ungroup()
data.org1 <- data.org1 %>%
  mutate(
    Year = as.factor(Year),
    Location = as.factor(Location),
    Yield = as.numeric(Yield),
    sowingdate.f = as.factor(SowingdateDOY),
    crm.f = as.factor(CRM_group))
data.org1 <- data.org1 %>%
  group_by(region)
```

```{r}
mod_rs3 <- lme(
  Yield ~  Location + SowingdateDOY + SowingdateDOY2 * crm.f +
    above_35 + cum_rain + cum_radn + Year,
  random = ~1 | Hybrid,
  weights = varIdent(form = ~1 | Location),
  data = data.org1)
```

```{r}
partial_r2 <- r2beta(mod_rs3, method = "nsj", data = data.org1)
print(mod_rs3)
```

```{r}
partial_r2 %>%
  mutate(Group = case_when(
    grepl("CRM_group|Hybrid", Effect) ~ "Genetics",
    grepl("SowingdateDOY|:CRM_group", Effect) ~ "Management",
    grepl("above_35|cum_rain|cum_radn|Year|Location", Effect) ~ "Environment",
    TRUE ~ "Other")) %>%
  group_by(Group) %>%
  summarise(Group_R2 = sum(Rsq)) %>%
  arrange(desc(Group_R2))

```



## Fitting one model per Region
previous model has one single slope for all regions son it cant quantify the effect of each factor by region.

```{r}
df_NE <- data.org1 %>%
  filter(region == "NorthEast")
summary(df_NE)
```


```{r}
mod_NE <- lme(
  Yield ~ Year +SowingdateDOY + SowingdateDOY^2 * crm.f +
            above_35 + cum_rain + cum_radn,
  random  = list(Hybrid = ~1),
  #weights = varIdent(form = ~1 | Location), 
  data = df_NE)
summary(mod_NE)
```
## Variance partitioning



```{r}
df_CE <- data.org1 %>%
  filter(region == "CenterEast")
summary(df_CE)
```

```{r}
mod_CE <- lme(
  Yield ~ Location + Year + SowingdateDOY + SowingdateDOY^2 * crm.f +
            above_35 + cum_rain + cum_radn,
  random  = list(Hybrid = ~1),
  weights = varIdent(form = ~1 | Location), 
  data = df_CE)
summary(mod_CE)
```

```{r}
plot(df_CE$Yield)
hist(df_CE$Yield)
```

##
##Correlation of weather variables with yield


```{r}
 climate_vars <- c("mean_mint","cum_radn","cum_rain",
                  "above_35", "april","may","june","july", "sowingdate_doy", "crm")

df_cor <- df %>%
  select(yield, all_of(climate_vars)) %>%
  drop_na()


cor_mat <- cor(df_cor, method = "pearson")


cor_melt <- melt(cor_mat, varnames = c("Variable1", "Variable2"), value.name = "Correlation")

# filtrar solo correlaciones con yield
cor_yield <- cor_melt %>% filter(Variable1 == "yield" & Variable2 != "yield")

# heatmap simple (ordenado)
p8 <- ggplot(cor_yield, aes(x = reorder(Variable2, Correlation), y = Correlation, fill = Correlation)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(x = "", y = "Pearson correlation with Yield (tn/ha)",
       title = "Correlation between climatic variables and corn yield") +
  theme_minimal(base_size = 13)
ggsave(plot = p8, "output/correlation.tiff", width = 8, height = 7 , unit = "in", dpi = 600,bg="white")
```


## model

##################################################################################

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(nlme)
library(MuMIn)
library(performance)
library(reshape2)   
library(tidyr)
library(tibble)
```


```{r}
df <- read.csv("input/Corn data organized1.csv", stringsAsFactors = FALSE)
df <- df %>% mutate(yield = as.numeric(yield)) %>% filter(!is.na(yield))

```

```{r}
climate_vars <- c("mean_mint","cum_radn","cum_rain","above_35","april","may","june","july")
```

```{r}
label_map <- c(
  mean_mint = "Avg min T",
  cum_radn  = "Cumulative radiation",
  cum_rain  = "Cumulative rain",
  above_35  = "Days with T >35",      
  april     = "April",
  may       = "May",
  june      = "June",
  july      = "July")
```

```{r}
df_cor <- df %>%
  select(yield, all_of(climate_vars)) %>%
  drop_na()

cor_mat <- cor(df_cor, method = "pearson")
cor_melt <- melt(cor_mat, varnames = c("Variable1", "Variable2"), value.name = "Correlation")
cor_yield <- cor_melt %>%
  filter(Variable1 == "yield" & Variable2 != "yield") %>%
  mutate(Variable2 = factor(Variable2, levels = Variable2[order(Correlation)]),
         label = label_map[as.character(Variable2)])
```

```{r}
correlation <- ggplot(cor_yield, aes(x = reorder(label, Correlation), y = Correlation, fill = Correlation)) +
  geom_col(width = 0.7, color = "gray30") +
  coord_flip() +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#4575B4", midpoint = 0,
                       guide = guide_colorbar(barwidth = 12, barheight = 0.6)) +
  geom_text(aes(label = sprintf("%.2f", Correlation)), 
            hjust = -0.2, size = 5, fontface = "bold", color = "black") +
  labs(x = NULL,
       y = "Pearson correlation with yield (t ha⁻¹)",
       title = "Correlation between climatic variables and corn yield") +
  theme_classic(base_size = 25) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    axis.text = element_text(size = 14, color = "black"),
    axis.title = element_text(size = 14, face = "bold"),
    legend.position = "none",
    plot.margin = margin(8, 50, 8, 8)) +
  expand_limits(y = max(cor_yield$Correlation) * 1.12) 
ggsave("output/correlation.tiff", plot = correlation,
       dpi = 600, width = 8, height = 6, units = "in", compression = "lzw", bg= "white")
```



##model

```{r}
vars_sel <- c("mean_mint","cum_radn","cum_rain","above_35",
              "april","may","june","july")

data_model <- df %>%
  select(yield, year, location, hybrid, all_of(vars_sel)) %>%
  drop_na()
```

```{r}
vars_ok <- vars_sel[sapply(data_model[vars_sel], function(x) sd(x, na.rm = TRUE) > 0)]

data_model <- data_model %>%
  mutate(across(all_of(vars_ok), ~ as.numeric(scale(.x)), .names = "z_{col}"))

z_vars <- paste0("z_", vars_ok)
```


```{r}
form_fixed <- as.formula(paste("yield ~", paste(z_vars, collapse = " + ")))

m_climate <- lme(
  fixed   = form_fixed,
  random  = ~ 1 | year/location/hybrid,
  data    = data_model,
  method  = "REML",
  control = nlme::lmeControl(msMaxIter = 200, msMaxEval = 200))

summary(m_climate)
```

```{r}
coef_rank <- summary(m_climate)$tTable %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  dplyr::filter(!grepl("(Intercept)", variable)) %>%
  mutate(abs_effect = abs(Value)) %>%
  arrange(desc(abs_effect)) %>%
  select(variable, Value, Std.Error, `t-value`, `p-value`, abs_effect)

print(coef_rank)
```


```{r}
library(tidyverse)
library(dplyr)
library(nlme)
library(MuMIn)
library(dplyr)
```

```{r}
df <- read.csv("input/Corn data organized1.csv", stringsAsFactors = FALSE)
```


```{r}
data_model <- df %>%
  select(yield, year, location, hybrid, mean_mint, above_35, cum_radn, cum_rain, april, may, june, july,
         august, sept)  %>% 
  mutate(yield= as.numeric(yield),
                above_35 = as.numeric(above_35))%>%
  drop_na()
```

```{r}
m_climate <- lme(
  fixed   = yield ~ mean_mint + above_35 + cum_rain + cum_radn + april + may +
    june + july + august + sept,
  random  = ~ 1 |location/hybrid,
  data    = data_model,
  method  = "REML",
  control = lmeControl(msMaxIter = 200, msMaxEval = 200))
summary(m_climate)
```

```{r}
r2_vals <- r.squaredGLMM(m_climate)
r2_vals
```


```{r}
coef_rank <- summary(m_climate)$tTable %>%
  as.data.frame() %>%
 rownames_to_column("variable") %>%
  filter(!grepl("(Intercept)", variable)) %>%
  mutate(abs_effect = abs(Value)) %>%
  arrange(desc(abs_effect)) %>%
  select(variable, Value, Std.Error, `t-value`, `p-value`, abs_effect)

print(coef_rank)

ggplot(coef_rank, aes(x = reorder(variable, abs_effect), y = abs_effect)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = NULL, y = "Absolute effect size (|β|)",
       title = "Relative importance of climate predictors (raw units)") +
  theme_classic(base_size = 16)
```

### all variables are in different units so I need to scale them
Z scores final graph


```{r}
df <- read.csv("input/Corn data organized2.csv", stringsAsFactors = FALSE)
data_model <- df %>%
  select(yield, year, location, hybrid, mean_mint, above_35, cum_radn.j,vpd, april, may, june, july,august, sept, precipitation) %>% 
  mutate(yield= as.numeric(yield),
                above_35 = as.numeric(above_35))%>%
  drop_na()
```

```{r}
data_model_z <- data_model %>%
  mutate(
    z_mean_mint = as.numeric(scale(mean_mint)),
    z_above_35  = as.numeric(scale(above_35)),
    z_precipitation  = as.numeric(scale(precipitation)),
    z_cum_radn  = as.numeric(scale(cum_radn.j)),
    z_vpd       = as.numeric(scale(vpd)),
    z_april     = as.numeric(scale(april)),
    z_may       = as.numeric(scale(may)),
    z_june      = as.numeric(scale(june)),
    z_july      = as.numeric(scale(july)),
    z_august    = as.numeric(scale(august)),
    z_sept      = as.numeric(scale(sept)))
```


```{r}
m_climate_z <- lme(
  fixed  = yield ~ z_mean_mint + z_above_35 + z_cum_radn + z_vpd + z_april +
    z_may  + z_july + z_august + z_sept,
  random = ~ 1 | location/hybrid,
  weights = varIdent(form = ~1 | location),  
  data   = data_model_z,
  method = "REML")
```

```{r}
qqnorm(residuals(m_climate_z))
qqline(residuals(m_climate_z), col = "red", lwd = 2)
#plot(m_climate_z$residuals)
```

```{r}
coef_rank_z <- summary(m_climate_z)$tTable %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  filter(variable != "(Intercept)") %>%
  mutate(abs_effect = abs(Value)) %>%
  arrange(desc(abs_effect)) %>%
  select(variable, Value, Std.Error, `t-value`, `p-value`, abs_effect)

print(coef_rank_z)
```

```{r}
coef_rank_z <- coef_rank_z %>%
  mutate(variable = str_replace(variable, "z_mean_mint", "Avg min T"),
         variable = str_replace(variable, "z_above_35", "Days T > 35°C"),
         variable = str_replace(variable, "z_cum_rain", "Cumulative rain"),
         variable = str_replace(variable, "z_cum_radn", "Cumulative radiation"),
         variable = str_replace(variable, "z_april", "Apr (mm)"),
         variable = str_replace(variable, "z_may", "May (mm)"),
         variable = str_replace(variable, "z_june", "Jun (mm)"),
         variable = str_replace(variable, "z_july", "Jul (mm)"),
         variable = str_replace(variable, "z_august", "Aug (mm)"),
         variable = str_replace(variable, "z_sept", "Sep (mm)"))
```


```{r}
standardized <-  ggplot(coef_rank_z, aes(x = reorder(variable, abs_effect), y = abs_effect, fill = abs_effect)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", abs_effect)), hjust = -0.2, size = 4.5, family = "sans") +
  scale_fill_gradient(low = "lightblue", high = "steelblue") +
  coord_flip() +
  labs(
    x = NULL,
    y = expression("Absolute standardized effect size (|"*beta*"|)"),
    title = "Relative importance of climate predictors") +
  theme_classic(base_size = 16) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "gray40"),
    panel.grid.major.x = element_line(color = "gray85", linetype = "dotted"),
    legend.position = "none",
    plot.margin = margin(10, 30, 10, 10)) +
  expand_limits(y = max(coef_rank_z$abs_effect) * 1.15)
ggsave(plot = standardized, "output/standardized.tiff",
        width = 8, height = 7, unit = "in", dpi = 600, bg = "white")
```




